---
title: "Extract Data from Seurat Object"
author: "Ruifeng Hu"
date: "`r Sys.Date()`"
description: "This notebook extracts data from a Seurat object and saves it in formats suitable for downstream analysis."
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

## Overview

This notebook extracts data from a Seurat object and saves it in formats suitable for downstream analysis.

## Usage

Run as an R script:

```
Rscript extract_SC_v5.R seurat_obj.rds output_directory cluster_column
```

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
library(jsonlite)
library(Matrix)
library(data.table)
library(Seurat)
library(tidyverse)
library(presto)
```

## Set Arguments

You can set these manually for interactive use:

```{r input-args}
# Set these manually for notebook use
seurat_obj_file <- "example_data/pmdbs_lee_obj_updated_SN.rds"  # Path to your Seurat object fil
output_dir <- "example_data/PMDBS_snRNAseq"  # Directory to save output files, usually the dataset name
cluster_col <- "cell_type"  # Column name in metadata for clustering information
```

## Create Output Directory

```{r create-output_dir}
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

## Load Seurat Object

```{r load-seurat-object}
cat("Loading RDS data...\n")
seurat_obj <- readRDS(seurat_obj_file)
if (!inherits(seurat_obj, "Seurat")) {
  stop("The provided file is not a valid Seurat object.")
}
capture.output(str(seurat_obj), file = paste0(output_dir, "/seurat_obj_structure.txt"))
```

## Check Seurat Object Structure
v5 Seurat objects have a different structure compared to v4, so we need to ensure the object contains the expected slots and layers.
```{r check-seurat-structure}
if (!"RNA" %in% names(seurat_obj@assays)) {
  stop("The Seurat object does not contain the 'RNA' assay.")
}
if (!"data" %in% names(seurat_obj@assays$RNA@layers)) {
  stop("The Seurat object does not contain the 'data' slot in the 'RNA' assay.")
}
if (!"meta.data" %in% slotNames(seurat_obj)) {
  stop("The Seurat object does not contain the 'meta.data' slot.")
}
if (!"umap" %in% names(seurat_obj@reductions)) {
  stop("The Seurat object does not contain the 'umap' reduction.")
}
if (!cluster_col %in% colnames(seurat_obj@meta.data)) {
  stop(paste("The cluster column", cluster_col, "does not exist in the metadata."))
}
```

## Extract and Save Metadata

```{r save-metadata}
cat("Saving metadata...\n")
metadata <- seurat_obj@meta.data
write.csv(metadata, file = paste0(output_dir, "/raw_metadata.csv"), row.names = TRUE)

cat("Saving metadata feature names...\n")
column_names <- colnames(metadata)
write_json(column_names, path = file.path(output_dir, "raw_metadata_columns.json"), pretty = TRUE)
```

## Save UMAP Embedding

```{r save-umap}
cat("Saving UMAP embedding...\n")
umap_embeddings <- seurat_obj@reductions$umap@cell.embeddings
colnames(umap_embeddings) <- toupper(colnames(umap_embeddings))
write.csv(umap_embeddings, file = paste0(output_dir, "/raw_umap_embeddings.csv"), row.names = TRUE)
```

## Extract and Save Normalized Counts

```{r save-normalized-counts}
cat("Saving normalized counts...\n")
normalized_counts <- seurat_obj[["RNA"]]@layers[["data"]] # Sparse matrix

gene_names <- dimnames(seurat_obj[["RNA"]]@features)[[1]]
cell_names <- dimnames(seurat_obj[["RNA"]]@cells)[[1]]

triplet <- summary(normalized_counts)
triplet$Gene <- gene_names[triplet$i]
triplet$Cell <- cell_names[triplet$j]

long_data <- triplet %>%
  select(Cell, Gene, Expression = x)

nonzero_data <- long_data[long_data$Expression > 0, ]
nonzero_data <- as.data.table(nonzero_data)
fwrite(nonzero_data, file = paste0(output_dir, "/raw_normalized_counts.csv"), row.names = FALSE)
```

---

```{r Done, echo=TRUE}
cat("Done! Data extraction is complete!
You can find the output files in the directory:", output_dir, "\n")
```